### 2014-10-27 core-lang.tex
Minor edits

We demonstrate our ideas using a replicated data store formalized as a
triple $\E;\Theta;\Sigma$, where $\E$ denotes the execution state,
$\Theta$ maps replicas to their local state, and $\Sigma$, the session
soup, is the set of client sessions interacting with the store.
Operational semantics given in Figure~\ref{sem:oper} succinctly
capture the behaviour of the store that is relevant to demonstrate the
operational significance of contract classification. For the sake of
clarity, we only consider a single replicated object of well-defined
type (for eg: a replicated object of type {\sf bank\_account}) in our
formalization.  Our semantics are parametric over the specification of
this replicated data type.  Figure~\ref{sem:oper} formalizes
replicated data type (RDT) specification as tuple
$(\tau,\Ops,\Ctrts)$, where $\tau$ is the data type, $\Ops$ maps
labels ($op$) of operations on $\tau$ to their definitions, while
$\Ctrts$ maps them to their consistency contracts ($\cv$). The language of
contracts is described in Fig.~\ref{lang:contract}. The definition of
an operation is expected to be a lambda expression in the server-side
language\footnote{Since a replicated object is maintained by a
server, we call the language in which replicated data type operations
are defined as \emph{server-side language}.}, although we do not
enforce this in our formalization. We intentionally leave the
server-side language underspecified to demonstrate that our ideas are
independent of the choice of the programming language used to
implement the replicated data type.

@@

An execution state $\E$ is a tuple $\Exec$ where $\EffSoup$ is a set
of \emph{effect}s , and $\visZ,\soZ \subseteq \EffSoup \times
\EffSoup$ are visibility and session order relations over effects,
respectively. An effect ($\eff$), modeled as a tuple $(s,i,op,v)$,
records the fact that $i^{th}$ action in session with {\sf SessID}
$s$, which is an operation $op$ on the replicated object, has been
successfully executed on some replica yielding a return value $v$.
Note that the combination of $s$ and $i$ uniquely identifies the
effect. Session order relation ($\soZ$) relates effects generated by
the same session. An effect $\eff = (s,i,op,v)$ is said to precede
another effect $\eff' = (s',i',op',v')$ in session order if and
only if $s'=s$ and $i'\ge i$. An effect generated at a replica becomes
\emph{available} at rest of the replicas eventually. The local state
of a replica $r$ (i.e., $\Theta(r)$) is a set of effects that are
currently available at $r$. An operation $op$ performed at replica $r$
can only witness the set of effects ($\Theta(r)$) visible at $r$. If we denote the
effect generated by the operation $op$ as $\eff_{op}$, then $\Theta(r)
\times \{\eff_{op}\} ~\subseteq~ \visZ$. Note that in presence of
multiple replicated objects, $\visZ$ only relates effects of
operations performed on a single object. For instance, a {\sf
withdraw} operation can witness (via $\visZ$) a {\sf deposit}
operation on the same bank account, but not a {\sf deposit} on a
different account. Often, in our formalization, we use $\visZ$ and
$\soZ$ binary relations to obtain a set of effects visible to a given
effect $\eff$, or set of effects that precede a given effect $\eff$ in
the session order. As a syntactic convenience, whenever $R$ is a
binary relation, we write $R(\eff)$ to denote the set of all $\eff'$
such that $(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$
to denote the set of all $\eff'$ such that $(\eff',\eff) \in R$.

Rule $\rulelabel{Comm}$ captures inter-replica communication. The
basic guarantee provided by the store is causal consistency. That is,
an effect ($\eff$) will become \emph{available} at a replica $r$ 
only after all the effects that causally precede $\eff$ become
available at $r$. Observe that this condition is applicable even for
the effects that originate at replica $r$. It is helpful to think of
such effects being stored in a \emph{write buffer} until causal
consistency condition is met, at which point they can be merged with
the local state ($\Theta(r)$). Nevertheless, it is important to note
that that enforcing causal consistency does not require inter-replica
coordination.

@@

\noindent Under the store configuration $\Theta$, the rule captures
the progress in execution (from $\E$ to $\E'$) due to the application
of operation $op$ to replica $r$ corresponding to a new effect $\eff$.
The rule first constructs a \emph{context} for the application from
the local state ($\Theta(r)$) of the replica, by
projecting\footnote{{\textsf{ctxt*}} is auxiliary function
\textsf{ctxt} extended straightforwardly to set of effects} relevant information from
effects in $\Theta(r)$. It then substitutes the definition
($\Ops(op)$) of the operation for its label ($op$), and relies on the
reduction relation ($\rdtredsto$) of the server-side language to
reduce the application $\Ops(op)(v,~ctxt)$ to a value $v'$.
Subsequently, the attributes of the new effect $\eff$ in the execution
state. In particular, we record that the effects in $\Theta(r)$ are
visible to $\eff$, and $\eff$ follows in session order ($\soZ$) every
effect in the same session $s$ with a smaller sequence number.

If the operation $op$ is highly-available (rule $\rulelabel{HA}$), we
simply apply the operation to any replica $r$. Since the store
provides causal visibility, highly available operations are
satisfiable under any replica. If the operation is sticky-available
(rule $\rulelabel{SA}$), the operation can be applied to a replica $r$
only if it already contains the effects of all the previous performed
in the same session. This guarantee can be satisfied by applying all
operations from the same session to the same logical copy of the
database. If such a logical copy is untenable, then the operation
might block. Since the store is assumed to converge eventually, the
blocked sticky-available operation guaranteed to unblock eventually.

>>>>

We demonstrate our ideas using a replicated data store formalized as a
triple $\E;\Theta;\Sigma$, where $\E$ denotes the execution state,
$\Theta$ maps replicas to their local state, and $\Sigma$, the session
soup, is the set of client sessions interacting with the store.
Operational semantics given in Figure~\ref{sem:oper} succinctly
capture the behaviour of the store that is relevant to demonstrate the
operational significance of contract classification. For the sake of
clarity, we only consider a single replicated object of well-defined
type (for eg: a replicated object of type \cf{BankAccount}) in our
formalization.  Our semantics are parametric over the specification of
this replicated data type.  Figure~\ref{sem:oper} formalizes
replicated data type (RDT) specification as tuple
$(\tau,\Ops,\Ctrts)$, where $\tau$ is the data type, $\Ops$ maps
labels ($op$) of operations on $\tau$ to their definitions, while
$\Ctrts$ maps them to their consistency contracts ($\cv$). The
definition of an operation is expected to be a lambda expression in
the server-side language\footnote{Since a replicated object is
maintained by a server, we call the language in which replicated data
type operations are defined as \emph{server-side language}.}, although
we do not enforce this in our formalization. We intentionally leave
the server-side language underspecified to demonstrate that our ideas
are independent of the choice of the programming language used to
implement the replicated data type.

@@

An execution state $\E$ is a tuple $\Exec$ where $\EffSoup$ is a set
of effects, and $\visZ,\soZ \subseteq \EffSoup \times
\EffSoup$ are visibility and session order relations over effects,
respectively. An effect ($\eff$), modeled as a tuple $(s,i,op,v)$,
records the fact that $i^{th}$ action in session with {\sf SessID}
$s$, which is an operation $op$ on the replicated object, has been
successfully executed on some replica yielding a return value $v$.
Note that the combination of $s$ and $i$ uniquely identifies the
effect. Session order relation ($\soZ$) relates effects generated by
the same session. An effect $\eff = (s,i,op,v)$ is said to precede
another effect $\eff' = (s',i',op',v')$ in session order if and only
if $s'=s$ and $i'\ge i$. An effect generated at a replica becomes
visible at rest of the replicas eventually. The local state of a
replica $r$ (i.e., $\Theta(r)$) is a set of effects that are currently
visible at $r$. An operation $op$ performed at replica $r$ can only
witness the set of effects ($\Theta(r)$) visible at $r$.  Note that
in presence of multiple replicated objects, $\visZ$ only relates
effects of operations performed on a single object. For instance, a
{\sf withdraw} operation can witness (via $\visZ$) a {\sf deposit}
operation on the same bank account, but not a {\sf deposit} on a
different account. Often, in our formalization, we use $\visZ$ and
$\soZ$ binary relations to obtain a set of effects visible to a given
effect $\eff$, or set of effects that precede a given effect $\eff$ in
the session order. As a syntactic convenience, whenever $R$ is a
binary relation, we write $R(\eff)$ to denote the set of all $\eff'$
such that $(\eff,\eff') \in R$.  Conversely, we write $R^{-1}(\eff)$
to denote the set of all $\eff'$ such that $(\eff',\eff) \in R$.

Rule $\rulelabel{Comm}$ captures inter-replica communication. The
basic guarantee provided by the store is causal visibility. That is,
an effect ($\eff$) is made visible at a replica $r$ only after all
the effects that causally precede $\eff$ are made visible at $r$.
Observe that this condition is applicable even for the effects that
originate at replica $r$. 
% It is helpful to think of effects being
% stored in a \emph{write buffer} until causal consistency condition is
% met, at which point they can be merged with the local state
% ($\Theta(r)$). 
It is important to note that that enforcing causal visibility does not
require any inter-replica coordination.

@@

\noindent Under the store configuration $\Theta$, the rule captures
the progress in execution (from $\E$ to $\E'$) due to the application
of operation $op$ to replica $r$ resulting in a new effect $\eff$.
The rule first constructs a \emph{context} for the application from
the local state ($\Theta(r)$) of the replica, by
projecting\footnote{{\textsf{ctxt*}} is auxiliary function
\textsf{ctxt} extended straightforwardly to set of effects} relevant
information from effects in $\Theta(r)$. It then substitutes the
definition ($\Ops(op)$) of the operation for its label ($op$), and
relies on the reduction relation ($\rdtredsto$) of the server-side
language to reduce the application $\Ops(op)(v,~ctxt)$ to a value
$v'$.  Subsequently, the the attributes of execution state, namely
$\EffSoup$, $\visZ$ and $\soZ$, are extended to incorporate
the new effect ($\eff$).

If the operation $op$ is highly-available (rule $\rulelabel{HA}$), we
simply apply the operation to any replica $r$. Since the store
provides causal visibility, highly available operations are
satisfiable under any replica. If the operation is sticky-available
(rule $\rulelabel{SA}$), the operation can be applied to a replica $r$
only if it already contains the effects of all the previous operations
from the same session. This guarantee can be satisfied by applying all
operations from the same session to the same logical copy of the
database. If such a logical copy is untenable, then the operation
might block. Since the store is assumed to converge eventually, the
blocked sticky-available operation guaranteed to unblock eventually.
